Well-quasi-orderings on word languages

### 2024-08-05 src/prefixes.tex
Add statements and proof scheme

% lang: en-US
\section{Prefixes and Suffixes}
\label{prefixes-suffixes:sec}

In this section, we study the well-quasi-ordering of languages under the prefix
relation. Let us first remark that the case of the suffix relation is
completely symmetric, and therefore that all our results will apply to the
suffix relation as well.

\begin{itemize}
    \item examples: $a^n$, $ab^n$, $b^na$. 
    \item the tree of prefixes and existence antichains
    \item decision procedure for regular languages
\end{itemize}

>>>>

% LTeX: language=en-GB 
\section{Prefixes and Suffixes}
\label{prefixes-suffixes:sec}

In this section, we study the well-quasi-ordering of languages under the prefix
relation. A word $u$ is a \intro{prefix} of a word $w$ if there exists a word
$v$ such that $w = uv$. We denote this relation by $u \intro*\prefleq w$.
Similarly, a word $u$ is a \intro{suffix} of a word $w$ if there exists a word
$v$ such that $w = vu$. We denote this relation by $u \intro*\suffleq w$.

Let us immediately remark that the map $u \mapsto u^R$ that reverses a word is
an order-bijection between $(X^*, \prefleq)$ and $(X^*, \suffleq)$. Therefore,
we will focus on the prefix relation in the following.

As mentioned in the introduction, the prefix relation is not a well-quasi-order
on words over a two-letter alphabet, as witnessed by the infinite
\kl{antichain} $\setof{a^nb}{n \in \Nat}$.

Let us order the words in $\Sigma^*$ by the \kl{prefix relation} to build an
infinite tree $T$. The root of the tree is the empty word $\varepsilon$, and
the children of a word $w$ are the words $wa$ for all $a \in \Sigma$. This tree
is finitely branching. Let $L \subseteq \Sigma^*$ be a language. Assume that
$L$ contains an infinite \kl{antichain} $A$ for the \kl{prefix relation}. In
particular, no two word of $A$ belong to the same branch of the tree $T$. Let
us prove the following: there exists an infinite branch $B$ of $T$ such that from
every point of the branch, one can reach a word of $A$ that is not in the
branch.

\begin{definition}
    An \intro{antichain branch} for a language $L$ is an infinite 
    branch $B$ of the tree $T$ such that from every point of the branch, 
    one can reach a word of $A$ that is not in the branch.
\end{definition}

\begin{lemma}
    Let $L \subseteq \Sigma^*$ be a language. Then, $L$ contains an infinite
    \kl{antichain} if and only if there exists an \kl{antichain branch} for $L$.
\end{lemma}

\begin{corollary}
    If $L$ is regular, then the existence of an infinite antichain is decidable.
\end{corollary}

\begin{theorem}
    A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the
    \kl{prefix relation} if and only if $L$ is a union of \kl{chains}.
\end{theorem}

\begin{corollary}
    Let $L \subseteq \Sigma^*$ be a language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists a finite $k \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $k$.
\end{corollary}
